---
title       : The Joys of Data Visualisation Using R
subtitle    : 
author      : Ewan Keith
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## What is R?

> * A statistical programming language.
> * Most importantly, it is:
   * Free!
   * Open Source
   * Under Constant Development

--- 

## Why Use R?

> * It is desgined with statistics and statisticians in mind
> * Provides a single platform for analysis and visualisation
> *	Massive range of visualisation options
> *	On the cutting edge/constantly updated 
> *	Range of dissemination options
> *	Shiny 


---

## Designed For Statistics

> * Basic plots are very easy to produce
> * also easily altered quickly
> * Well suited to fast, exploratory visualisation


---
## Basic Plots are easy

```{r echo = FALSE}
# prep log-normal and iris data
x <- rlnorm(500)
data("iris")
```

```{r}
plot(iris$Petal.Length, iris$Petal.Width)
```

---
## Brief aside on Flower data



---
## Basic Plots are easy
```{r}
hist(x)
```


---
## Plots easily altered
```{r}
x2 <- x[x < 10]
hist(x2)
```


---
## Plots easily altered
```{r}
hist(log(x))
```


---
## Common Statistical Tasks Are Easily Handled

> * Basic Statistical tasks are considered and easy to carry out.
    * which they may not be in other programming languages.
    
> * Major reason to consider R over alternatives like Python or Matlab.

> * Similar to SAS in this regard.


---
## Common Statistical Tasks Are Easily Handled

```{r echo=FALSE}
# load the lattice package for xyplot()
library(lattice)
```

```{r}
plot(iris$Petal.Length, iris$Petal.Width)
```


---
## Common Statistical Tasks Are Easily Handled

```{r}
xyplot(Petal.Width ~ Petal.Length | Species,data = iris)
```


---
## Common Statistical Tasks Are Easily Handled
```{r}
pairs(iris[,1:4])
```


---
## Common Statistical Tasks Are Easily Handled

* Model Checking can be time consuming given the number of assumptions that need checked

```{r echo = FALSE}
regression_model <- lm(Petal.Width ~ Petal.Length, data = iris)
plot(iris$Petal.Width ~ iris$Petal.Length)
abline(regression_model)
```


---
## Common Statistical Tasks Are Easily Handled

```{r eval = FALSE}
plot(regression_model)
```

```{r echo = FALSE}
par(mfrow=c(2,2))
plot(regression_model)
```


---
##	Publication Quality Plots are Easy

> * Once a basic plot has been decided on it is easy to produce high quality versions for publication.
> * Highly customisable and available in a range of formats.


---
## Publication Quality Plots are Easy

```{r echo = FALSE}

library(ggplot2)
library(ggthemes)

ggplot(iris, aes(Petal.Width, fill=Species)) +
  geom_density(alpha=0.5) + scale_color_fivethirtyeight() +
 theme_fivethirtyeight() + ggtitle("Petal Width Distribution")
```


---
## Publication Quality Plots are Easy

```{r echo = FALSE}
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_point() + 
    geom_smooth(method = lm) + scale_color_fivethirtyeight() +
  theme_fivethirtyeight()
```


---
## Designed for Statistics

> *	All of the above can be done by any most other programming languages
> * But not as easily, as neatly, or as quickly


---
## An Integrated Analysis & Visualisation Platform

> * Allows you to easily move from analysis to visualisation.
> * Keeps the workflow clean and can be stored in a single file.
> * Allows for easy visualisation of complex analyses/models.
> * Easy to replicate/extend plots for new data


---
## An Integrated Analysis & Visualisation Platform

```{r echo = FALSE}
# example taken from http://stackoverflow.com/questions/18424133/r-forecast-function-not-picking-up-seasonality
sw<-c(2280, 1754, 1667, 1359, 1285, 1379, 2166, 1053, 1076, 1149, 1277, 1577, 1639, 1719, 1592, 2306, 3075, 2897, 1875, 1966, 2927, 3528, 2948, 2890, 3947, 3913, 3885, 4148, 5293, 5752, 6001, 7719, 5512, 6782, 6320, 6425, 6406, 7237, 8655, 9269, 12447, 13470, 13469, 13949, 17753, 17653, 14531, 14496, 13643, 12652, 12665, 10629, 8962, 8198, 6833, 5027, 4407, 4449, 4399, 5896, 6589, 3786, 4386, 4847, 5597, 5407, 4800, 7803, 9255, 10423, 5523, 8121, 6944, 8434, 9847, 9292, 9794, 10195, 10124, 11310, 12245, 12798, 14611, 15402, 13532, 16154, 15101, 14755, 17139, 16475, 19935, 19980, 25173, 28568, 27839, 28991, 27073, 29615, 25849, 27910, 27067, 21303, 20544, 15188, 13706, 9277, 10815, 7228, 4608, 4409, 9866, 8471, 8223, 6445, 6641, 6833, 11421, 8945, 8127, 10380, 12005, 13272, 9431, 12144, 14934, 14052, 11712, 14888, 15824, 17275, 18067, 19839, 21192, 22763, 22976, 23721, 22681, 20131, 19965, 20539, 19517, 22022, 23076, 30574, 40247, 43111, 39577, 40724, 44982, 44388, 46372, 43153, 36821, 32258, 31256, 27153, 23180, 18252, 16381, 13220, 12500, 10727, 9636, 8892, 8644, 9482, 9170, 10937, 12299, 15781, 11477, 16524, 16752, 18072, 14776, 13388, 18056, 19815, 21263, 22046, 26415, 24247, 25403, 30058, 26331, 32533, 31891, 35973, 27558, 24554, 25692, 25955, 24284, 24930, 28354, 34840, 40055, 42099, 42768, 48279, 50086, 56466, 42244, 51451, 44583, 39091, 33391, 29452, 25533)

swts <- ts(sw, frequency=52, start=c(2006,30))

swfc <- forecast(swts,h=52)

lambda <- BoxCox.lambda(sw)
swfc <- forecast(swts,h=52, lambda = lambda, robust = TRUE)
plot(swfc)
```


---
## An Integrated Analysis & Visualisation Platform

```{r echo = FALSE}
library(lavaan)

# Example 5.8 from mplus user guide:
Data <- read.table("http://www.statmodel.com/usersguide/chap5/ex5.8.dat")
names(Data) <- c(paste("y", 1:6, sep = ""), paste("x", 1:3, sep = ""))

# Model:
model.Lavaan <- "f1 =~ y1 + y2 + y3\nf2 =~ y4 + y5 + y6\nf1 + f2 ~ x1 + x2 + x3 "

# Run Lavaan:
fit <- lavaan:::cfa(model.Lavaan, data = Data, std.lv = TRUE)

# Plot path diagram:
library(semPlot)
semPaths(fit, title = FALSE, curvePivot = TRUE)
```


---
## An Integrated Analysis & Visualisation Platform

```{r echo = FALSE}
# NOTE this package requires a seperate installation of JAGS software, will not be reproducable without this
library(BayesianFirstAid)

d <- read.csv("2d4d_hone_2012.csv")
fit_male <- bayes.cor.test( ~ ratio_2d4d + grip_kg, data = d[ d$sex == "male",])
plot(fit_male)
```


---
## Massive Range of Visualisation Options

> * 'Basic' Visualisations extremely well integrated
> * But much greater flexibility is offered than in most other software packages
> * Wide range of domain specific visualisation supported also (e.g. GIS)


---
## 'Basic' Charts

```{r echo = FALSE}
library(treemap)
data(GNI2010)
treemap(GNI2010,
index=c("continent", "iso3"),
vSize="population",
vColor="GNI",
type="value")
```


---
## 'Basic' Charts
```{r echo = FALSE}
data("attitude")
library(reshape2)
qplot(x=Var1, y=Var2, data=melt(cor(attitude)), fill=value, geom="tile")
```


---
## 'Basic' Charts
```{r echo = FALSE}
nodes <- read.csv("Dataset1-Media-Example-NODES.csv", header=T, as.is=T)
links <- read.csv("Dataset1-Media-Example-EDGES.csv", header=T, as.is=T)
suppressMessages(
library(igraph)
)
net <- graph.data.frame(links, nodes, directed=T)
plot(net, edge.arrow.size=.4,vertex.label=NA)

```


---
## Specific Subject Areas

> * R also has very strong capabilities in topic specific areas of visualisation
> * e.g. Geographical Information Systems (GIS)
>   * Again R excels here as a single analysis/visualisation environment.


---
## GIS Capabilities

```{r echo=FALSE}
suppressMessages(
library(ggmap)
)

bristol.drugs <- read.csv("Bristol Drug Crimes.csv")
load("Bristol map.rda")

ggmap(bristol.map, extent = "device", legend = "topleft", extend = "panel") +
    geom_point(
      aes(x = Longitude, y = Latitude),
      data = bristol.drugs, na.rm=T) +
    ggtitle("Drug Crimes in Bristol\n from Feb 2014 - Feb 2015") +
    theme(plot.title = element_text(lineheight=.8, face="bold"))

```


---
## GIS Capabilities
```{r echo = FALSE}
ggmap(bristol.map, extent = "device", legend = "topleft", extend = "panel") +
    stat_density2d(
      aes(x = Longitude, y = Latitude, fill = ..level..),  alpha = .3,
      data = bristol.drugs,
      geom = "polygon", na.rm=T) +
    ggtitle("Drug Crimes in Bristol\n from Feb 2014 - Feb 2015") +
    theme(plot.title = element_text(lineheight=.8, face="bold"))

```

---
## GIS Capabilities

* Commute Example


---
## On Cutting Edge

> * Constantly developed so new technology often encorporated
> * Has powerful support to the D3 Javascript plotting library
> * Google motion plots easily handled (if DII allowed Flash.)
> * Ever increasing number of options for interactive plotting (plot.ly went open source in November for example)

> * But just because this exists doesn't mean it always plays well with other packages. This includes slides...

```{r eval = FALSE, echo = FALSE}

## examples below to show outside slides, neccesary packages not all on CRAN so may be unavailable on the VM

# library(threejs)
# scatterplot3js(iris[,2:4], color=rainbow(3)[iris$Species], labels=iris$Species, renderer="canvas")

# library(rCharts)
# nPlot(Freq ~ Hair, group = 'Eye', data = hair_eye_male, type = 'multiBarChart')

# library(rchess)
# plot(Chess$new())


```

---
## Range of Dissemination Options

> * Not Limited to 'copy and pasting' images into word documents
> * R can be used to produce, fully functional:
> * Word Documents
> * Web Pages
> * PDF Documents
> * and slide shows!


---
## Examples of Dissemination

> * DEA Supplementary Web Page
> * Treemap Documentation
> * These slides
> * Automatically updating report


---
## R Shiny

> * Tool for producing interactive data applications
> * Extremely Powerful and Versatile
> * Can be hosted on a web server
> * Only requires knowledge of R code
> * Also completely Free

```{r echo = FALSE}
# examples
# as visualisation of official numbers - US Labour Force 
# learning tool - SS Reg
# For exploring any complicated dataset - football
# shiny stan - full blown analytical software

```


---
## So What?

> * R is currently available on the VM
> * Introductory training in R is currently being organised by the department
> * Working group set up recently to iron out any difficulties using R in MOD and look into extending functionality.


---
## Any Questions?









